% !TeX root = ../main.tex
\section{The Five Gates}
\label{sec:gates}

\subsection{Format gate $f_1$}

Let the structural specification be a formal language $L(G)$ (context-free
grammar, JSON-Schema compilation, or regular constraint set). Define
\[
f_1(y)=\frac{d_{\mathrm{edit}}(y,L(G))}{|y|},\qquad
d_{\mathrm{edit}}(y,L)=\min_{z\in L} d_{\mathrm{Lev}}(y,z).
\]
For CFGs this is computable exactly by error-correcting Earley/CYK parsing
\cite{aho1972} in $O(|G|^2|y|^3)$; for regular $L$, by product-automaton
shortest path in $O(|A||y|)$. The gate is deterministic with zero estimation
error, and $f_1(y)=0\iff y\in L(G)$. No Wasserstein projection is needed; the
v1 operator $\mathrm{Proj}_{P_{\mathrm{form}}}$ is replaced by an exact
combinatorial minimization with a known algorithm.

\subsection{Fact gate $f_2$}

Decompose $y$ into claims $c_1,\dots,c_m$; retrieve evidence $E_j$ from a
verified knowledge base $\mathcal K$; let $s_j$ be the calibrated entailment
probability from an NLI model (isotonic or Platt calibration on held-out data;
ECE reported). Per P3,
\[
f_2(y)=1-\min_{1\le j\le m}s_j,
\]
with a mandatory \emph{abstention state}: when retrieval confidence for a
claim falls below a floor, the output is routed to abstention rather than
scored---an unverifiable claim must not receive a numerical pass. (This
removes the v1 ambiguity in which kernel-smoothing of a reference measure
silently interpolated over missing knowledge.) Optionally, against a reference
sample from $\mathcal K$ one may use the Sinkhorn divergence
$S_\varepsilon(\hat\mu_y,\hat\nu_{\mathrm{fact}})$, which metrizes weak
convergence with dimension-robust sample complexity $O(n^{-1/2})$
\cite{genevay2019}; plug-in $W_2$ degrades as $O(n^{-1/d})$ \cite{fournier2015}
and is therefore not used anywhere in this paper.

\subsection{Logic gate $f_3$: weighted sheaf Laplacian, unified energy,
relax-and-certify}
\label{sec:logic-gate}

\paragraph{Claim graph.} $G=(V,E)$: vertices are claims; an edge $(u,v)$ with
relation type (entail / equiv / contradict) is added when the NLI model flags
it, carrying its calibrated confidence $w_e\in[0,1]$.

\paragraph{Weighted sheaf.} Attach to $G$ a cellular sheaf $\mathcal F$
\cite{hansen2019}: stalks $\mathcal F(v)=\mathbb R^k$ (truth coordinates of
claim $v$); restriction maps encode the relation (identity for entailment and
equivalence, the negation lift $t\mapsto 1-t$ for contradiction). With
coboundary $\delta:C^0\to C^1$ and $W=\mathrm{diag}(w_e)$, the weighted sheaf
Laplacian is $L_{\mathcal F}(w)=\delta^\top W\delta$, and the raw score of an
assignment $x$ is $\langle x,L_{\mathcal F}(w)x\rangle^{1/2}$: a
finite-dimensional Hodge-type quantity, computable in $O(|V|^3)$ worst case
and by sparse solvers in practice. Global sections
$\ker\delta=H^0(G;\mathcal F)$ are exactly the edge-consistent assignments, so
the score measures the obstruction to gluing the asserted claims into a global
section---the v1 intuition, now on a specified finite object.

\begin{proposition}[Lipschitz dependence on NLI confidences]
\label{prop:lipschitz}
If $\|x_v\|\le R$ and restriction maps have operator norm at most $1$, then
for confidence vectors $w,w'$,
\[
\bigl|\langle x,L_{\mathcal F}(w)x\rangle-\langle x,L_{\mathcal F}(w')x\rangle\bigr|
\le 4R^2\|w-w'\|_1 .
\]
\end{proposition}
\begin{proof}
The energy is $\sum_e w_e\|(\delta x)_e\|^2$ with $\|(\delta x)_e\|\le 2R$;
subtract termwise.
\end{proof}

An out-of-distribution NLI perturbation therefore produces a bounded,
quantified score perturbation, which composes with margin stability
(Prop.~\ref{prop:margin}); ``garbage in, garbage out'' is replaced by a
sensitivity bound. The bound is verified numerically in the reference
implementation.

\paragraph{Unified logic energy.} Propositional constraints enter as sheaf
data: lift Boolean assignments to $[0,1]$ stalks on the factor graph of
extracted constraints, with restriction maps encoding literal polarity; let
$X$ be the linear-relaxation polytope of the clauses. Define the (pairwise)
relaxed energy
\[
E_{\mathrm{pair}}(y)=\min_{\substack{x\in X\\ x\equiv 1 \text{ on asserted claims}}}
\langle x, L_{\mathcal F}(w)\, x\rangle^{1/2}.
\]
The objective is a convex piecewise quadratic; projected gradient or
interior-point methods find the global minimum. When a numeric extraction
channel is available (equation chains with value flow), the same construction
on affine restriction data yields a cycle-holonomy energy
$E_{\mathrm{num}}(y)$: on a spanning tree the asserted value deltas propagate
potentials, and each non-tree edge contributes its weighted closure gap
(length-invariant and localized at the closing edge).

\begin{definition}[Channel aggregation; max-form $f_3$]
\label{def:maxform}
With channel thresholds $\tau_{\mathrm{pair}},\tau_{\mathrm{num}}$ calibrated
on clean data (split conformal per channel),
\[
f_3(y)=\max\!\Bigl(\tfrac{E_{\mathrm{pair}}(y)}{\tau_{\mathrm{pair}}},\;
\tfrac{E_{\mathrm{num}}(y)}{\tau_{\mathrm{num}}}\Bigr).
\]
The aggregation form is forced by the same dilution phenomenon as
Prop.~\ref{prop:dilution} one level down: additive (quadrature) combination
lets pairwise-NLI noise on ordinary text drown the compositional channel. In
the field study of \S\ref{sec:field}, the quadrature form scores AUROC
$0.586$ on compositional violations and the max-form $0.990$ \emph{from the
same channel values}. Per-channel validity composes by the union bound
(Theorem~\ref{thm:familywise} applied channel-wise).
\end{definition}

\begin{proposition}[Relaxation gap]
\label{prop:gap}
(i) If the asserted constraint system is Boolean-satisfiable then
$E_{\mathrm{pair}}(y)=0$. (ii) The converse fails in general: LP/SDP
integrality-gap instances (e.g.\ MAX-2-SAT \cite{goemans1995}) yield
fractional zeros for unsatisfiable systems.
\end{proposition}
\begin{proof}[Proof sketch]
(i) A Boolean satisfying assignment lies in $X$ and has zero residual on every
edge by construction of the restriction maps. (ii) Standard gap instances
embed directly.
\end{proof}

\begin{proposition}[No exact polynomial unification]
\label{prop:no-poly}
A polynomial-time computable continuous score $s$ with $s(y)=0$ iff the
asserted propositional system is Boolean-satisfiable would decide SAT in
polynomial time. Hence, unless $\mathrm P=\mathrm{NP}$, every polynomial
unified score is incomplete on the hard fragment.
\end{proposition}

\begin{proposition}[Structural equivalence on assertion-only acyclic graphs]
\label{prop:struct-equiv}
Let $G$ be a claim graph in which every vertex is asserted (clamped to $1$)
and whose edges are entailments and contradictions with weights $w_e$; let
$C(G)$ denote its contradiction edge set. Then
\[
E_{\mathrm{pair}}(y)^2=\sum_{e\in C(G)}w_e ,
\]
i.e.\ the relaxed unified energy is a strictly monotone function of the
weighted contradiction sum (WSUM) and induces identical rankings (identical
AUROC and identical false-positive rate at every operating point).
\end{proposition}
\begin{proof}
With every vertex clamped there is no free variable. Entailment residuals
$\max(0,x_u-x_v)$ vanish at $x\equiv 1$, and each contradiction residual is
$x_u+x_v-1=1$; the constrained minimum is attained at the unique feasible
point.
\end{proof}

\begin{remark}[Where the sheaf can add capability]
\label{rem:boundary}
Prop.~\ref{prop:struct-equiv} delimits the tool's value: any detection
separation between the sheaf energy and weighted counting requires
\emph{structure}---free (non-asserted) vertices whose values the global
minimization can reconcile, cycles in the numeric/affine view whose holonomy
no single edge witnesses, or the certificate layer. The field study of
\S\ref{sec:field} confirms both directions of this boundary empirically.
\end{remark}

\paragraph{Relax-and-certify.} Consequently a SAT/SMT certifier is
retained---not as a parallel gate but as a \emph{certification layer}: it
verifies (rounds) near-zero relaxed optima and, on UNSAT, returns a
deletion-minimal unsatisfiable core whose size is an exact, interpretable
severity overriding a passing continuous score. The soft/hard dichotomy of
earlier drafts dissolves into the standard relax-and-certify hierarchy of
modern optimization. As a diagnostic, the edges carrying the largest weighted
residual energy localize the inconsistency; localization is evaluated in
\S\ref{sec:field}. A lattice-valued unification via the Tarski Laplacian
\cite{ghrist2022} is recorded as Open Problem~\ref{open:tarski}.

\subsection{Alignment gate $f_4$}

$f_4(y)=\sigma^{-1}(\hat p_{\mathrm{unsafe}}(y))$: the logit of a safety
classifier calibrated on held-out labels (ECE reported), composed with the
conformal layer of \S\ref{sec:calibration} so that a bounded false-rejection
rate on safe outputs holds as a theorem. \emph{Computational independence}
(structural requirement): $f_4$ operates on the raw text $y$ and does not
factor through $D$, the claim graph, or format parsing; a format-corrupted
output therefore cannot even computationally block its own safety evaluation.
Combined with the veto of \S\ref{sec:veto}, the masking channel is closed at
the dataflow level as well as the decision level. (The v1 formulation---$W_2$
projection onto a ``compact convex set of safe behaviors''---is dropped: the
set was never constructed and metric projection in positively curved
Wasserstein space is not well-posed.)

\subsection{Intent gate $f_5$}

\begin{proposition}[MMD dilution]
\label{prop:dilution}
Let $\hat\mu,\hat\mu'$ be empirical measures on $m$ units differing in exactly
one unit, with kernel bound $k\le K$. Then
$\MMD_k(\hat\mu,\hat\mu')\le 2\sqrt K/m$.
\end{proposition}
\begin{proof}
$\psi(\hat\mu)-\psi(\hat\mu')=\frac1m(\phi(e_j)-\phi(e'_j))$ and
$\|\phi(\cdot)\|\le\sqrt K$.
\end{proof}

A fatal single-unit violation (one inserted negation) moves an aggregate-MMD
score by $O(1/m)$---vanishing in long outputs. Hence, per P3,
\[
f_5(y)=\max\bigl(f_5^{\mathrm{hard}}(y),\,f_5^{\mathrm{soft}}(y)\bigr).
\]
\emph{Hard component (exact).} Compile the instruction into programmatically
checkable constraints $\{\kappa_\ell\}$ (length, inclusion/exclusion,
ordering---the IFEval regime \cite{zhou2023});
$f_5^{\mathrm{hard}}=\max_\ell \mathbf 1[\kappa_\ell\text{ violated}]$. A
single negation that flips a checkable constraint scores $1$ regardless of
$m$. \emph{Soft component (worst-unit, directed Hausdorff in feature space).}
\[
f_5^{\mathrm{soft}}(y)=\max_j\;\min_{r\in\mathrm{supp}(\nu_{\mathrm{intent}})}
\bigl\|\phi(E(u_j))-\phi(E(r))\bigr\|_{\mathcal H_k},
\]
where $\nu_{\mathrm{intent}}$ is the reference measure of the task objective.

\begin{proposition}[Unit-level sensitivity]
\label{prop:unit-sens}
If one unit lies at feature distance at least $\Delta$ from the entire intent
reference set, then $f_5^{\mathrm{soft}}\ge\Delta$, independently of $m$.
\end{proposition}

The kernel mean embedding $\psi(\mu)=\int\phi\,d\mu$ is isometric with respect
to MMD by construction, and MMD is a metric for characteristic $k$
\cite{gretton2012}. This replaces the v1 claim of an isometric RKHS embedding
of the $W_2$ geometry, which is impossible for $d\ge 3$ \cite{andoni2018}.
Three honesty notes: (i) cross-encoder relevance scores are generally not
positive semidefinite and may be used as raw scores, not kernels---conformal
calibration is agnostic to the score's provenance; (ii) the $O(n^{-1/2})$ MMD
concentration bound applies to the mean-embedding form; the max-form relies on
conformal calibration for validity and on \S\ref{sec:protocol} for empirical
power; (iii) aggregate MMD is retained solely as a drift-monitoring feature
(\S\ref{sec:regime3}), where distribution-level sensitivity is exactly what is
wanted.
