% !TeX root = ../main.tex
\section{Calibration: Finite-Sample Guarantees Under Three Regimes}
\label{sec:calibration}

\subsection{The guarantee is instrument-inclusive}

\begin{proposition}[Composite conformal validity]
\label{prop:composite}
Fix the instrument $I$ before calibration and let $f_i=g_i\circ I$. If
calibration and test outputs are exchangeable as raw texts (A3), then
Theorem~\ref{thm:split} below holds verbatim for the composite score.
\end{proposition}
\begin{proof}
Exchangeability is preserved under a fixed measurable map applied identically
to every element of the sequence; the rank argument of
Theorem~\ref{thm:split} then applies to the composite scores unchanged.
\end{proof}

Claim-extraction truncation and out-of-distribution NLI behavior therefore
cannot inflate the false-rejection rate: instrument noise enters calibration
and test exchangeably. What instrument error does affect is detection power
(Type-II), quantified in \S\ref{sec:power} and audited in
\S\ref{sec:protocol}.

\subsection{Regime I --- exchangeable deployment: split conformal}

With a calibration set of $n_i$ outputs certified valid on dimension $i$, set
$\tau_i$ to the $\lceil(n_i+1)(1-\alpha_i)\rceil$-th smallest calibration
score.

\begin{theorem}[Distribution-free false-rejection control]
\label{thm:split}
Under (A3), for a new valid output $y$, $\mathbb P(\rho_i(y)>1)\le\alpha_i$;
if scores are almost surely distinct, also
$\mathbb P(\rho_i(y)>1)\ge\alpha_i-\frac1{n_i+1}$.
\end{theorem}
\begin{proof}
By exchangeability, the rank of $f_i(y)$ among the $n_i+1$ scores is uniform
on $\{1,\dots,n_i+1\}$ (ties only help the upper bound). Exceedance of
$\tau_i$ requires the rank to exceed $\lceil(n_i+1)(1-\alpha_i)\rceil$, an
event of probability at most $\alpha_i$; the lower bound follows from the same
rank computation without ties.
\end{proof}

This is precisely the property v1's ``EVT conjecture'' asked for, obtained as
a three-line finite-sample theorem with no distributional assumptions.
(Extreme-value refinements remain optional; note that all gate scores here are
bounded, placing them in the Weibull, not Fr\'echet, max-domain of attraction
\cite{coles2001}---any EVT use must carry domain diagnostics.) The
field-study coverage audit (\S\ref{sec:field-e2}) verifies the exceedance
bound on real chains ($0.095\le\alpha=0.10$) and shows that the
coverage-\emph{distribution} refinement $\mathrm{Beta}(m,n+1-m)$ additionally
requires (near-)continuous scores: with $39.7\%$ tied scores it is rejected
while the bound itself holds; randomized tie-breaking or the tie-safe
statement should be used with discrete score atoms.

\begin{theorem}[Family-wise validity, safety-first]
\label{thm:familywise}
Choosing $\alpha_i=\alpha/5$ yields
$\mathbb P(\exists i:\rho_i(y)>1\mid y\text{ fully valid})\le\alpha$ by the
union bound. Where a fixed-sequence procedure is used to recover power under
dependence, gates are tested safety-first ($f_4$ at full level $\alpha$ before
all others), aligning power allocation with the veto semantics of
\S\ref{sec:veto}.
\end{theorem}

\subsection{Regime II --- estimable covariate shift: weighted conformal}

When the shift is estimable (e.g.\ a changed prompt mix), weighted conformal
prediction \cite{tibshirani2019} restores the per-instance guarantee with
likelihood-ratio weights $\hat w(x)$; weight-estimation error enters as an
explicit total-variation correction. (Online density-ratio estimation as in
\cite{cofact2026} is a compatible single-gate instance.)

\subsection{Regime III --- arbitrary/adversarial drift: adaptive conformal
inference}
\label{sec:regime3}

Deployment default: the online update of Gibbs and Cand\`es \cite{gibbs2021},
$\alpha_{t+1}=\alpha_t+\gamma(\alpha-\mathrm{err}_t)$, with the conformal
quantile taken at level $1-\alpha_t$.

\begin{theorem}[{\cite{gibbs2021}}, restated]
\label{thm:aci}
For arbitrary---including adversarially chosen---sequences of distributions,
\[
\Bigl|\tfrac1T\textstyle\sum_{t\le T}\mathrm{err}_t-\alpha\Bigr|
\le\frac{\max(\alpha_1,1-\alpha_1)+\gamma}{\gamma T}\to 0 .
\]
\end{theorem}

Long-run false-rejection frequency converges to nominal $\alpha$ with no
exchangeability or stationarity assumption; the guarantee does not lapse
between recalibrations because adaptation is per-step---there is no ``blind
window'' at the level of long-run frequency. (A3) is thereby demoted from a
load-bearing assumption to the condition under which the stronger per-instance
bound of Theorem~\ref{thm:split} additionally holds.

\paragraph{Monitoring and fail-safe posture.} Drift monitoring uses conformal
test martingales / e-processes \cite{vovk2021}: anytime-valid, no scheduled
testing lag, false-alarm control at every stopping time; the monitored
features include the aggregate-MMD statistic of \S\ref{sec:gates}. Upon alarm
and until re-validation, thresholds are tightened monotonically and low-margin
outputs are routed to abstention or human review.

\begin{proposition}[Safe-side degradation]
\label{prop:safeside}
Decreasing $\tau_i$ enlarges every rejection region and shrinks none; during
the alarm state the probability of a false accept on any dimension is
non-increasing relative to the pre-alarm configuration.
\end{proposition}
\begin{proof}
$\{f_i>\tau_i\}$ is decreasing in $\tau_i$ under set inclusion.
\end{proof}

What is sacrificed under drift is throughput (more abstentions), never the
safety direction of the error.

\subsection{Power accounting}
\label{sec:power}

\begin{proposition}[Detection power lower bound]
\label{prop:power}
If an inserted violation survives claim decomposition with probability at
least $1-\varepsilon_D$, is flagged as an edge by NLI with probability at
least $1-\varepsilon_{\mathrm{NLI}}$ given survival, and, once in the graph,
drives $\rho_3>1$ with probability at least $\pi_3$, then
\[
\mathbb P(\text{detected})\ge(1-\varepsilon_D)(1-\varepsilon_{\mathrm{NLI}})\,\pi_3 .
\]
\end{proposition}

$\varepsilon_D$ and $\varepsilon_{\mathrm{NLI}}$ are estimated on the audit
benchmarks of \S\ref{sec:protocol} with Clopper--Pearson intervals and enter
the reported power certificate. \emph{Abstention rule:} if extraction coverage
(fraction of tokens attributable to extracted claims) falls below a floor
$c_{\min}$, the output is routed to abstention.

\subsection{Reporting requirements}
\label{sec:reporting}

For every deployment: $n_i$; empirical coverage on held-out valid data with
Clopper--Pearson intervals; ECE for every internal classifier ($f_2$'s NLI,
$f_4$'s safety classifier); the e-process monitor state; and the regime
(I/II/III) under which each guarantee is claimed.
