% !TeX root = ../main.tex
\section{Pre-Registered Empirical Protocol}
\label{sec:protocol}

The protocol below was registered, and frozen, before any field data were
collected. Status markers record what the M1 field study has executed; the
remainder is the active experimental agenda.

\textbf{(1) Datasets and gate mapping.} \emph{Format:} structured-output
benchmarks plus synthetic grammar corruption at controlled edit distances
(ground-truth $f_1$). \emph{Fact:} TruthfulQA \cite{lin2022} and FEVER-style verification
with a frozen retrieval index. \emph{Logic:} GSM8K/MATH step-consistency with
adversarial contradiction insertion at known graph locations and controlled
distance $k$. \emph{[Executed for GSM8K in \S\ref{sec:field}; MATH pending an
audited equation extractor.]} \emph{Alignment:} human-adjudicated safety
benchmarks. \emph{Intent:} IFEval-style verifiable-constraint suites.

\textbf{(2) Primary metrics.} Per-gate AUROC and FPR at 95\% TPR; empirical
conformal coverage versus nominal $\alpha$ with Clopper--Pearson intervals (a
direct field test of Theorem~\ref{thm:split}) \emph{[executed:
\S\ref{sec:field-e2}]}; ACI long-run error tracking under injected drift
(Theorem~\ref{thm:aci}); ECE of internal classifiers \emph{[executed for the
NLI instrument]}; end-to-end $\dec$ accuracy against human triage.

\textbf{(3) Ablations with pre-registered predictions.}
(a) MMD vs.\ Sinkhorn vs.\ plug-in $W_2$: plug-in $W_2$ degrades with
embedding dimension per its $n^{-1/d}$ rate; the others do not.
(b) \emph{Sheaf vs.\ counting vs.\ SAT-only on real chains---the empirical
linchpin.} \emph{[Executed: \S\ref{sec:field}. Success criteria met on the
compositional cell (\(\Delta\)AUROC $+0.489$, Holm $p=0.0018$, \textsc{Wsum}
clause satisfied); the distance-collapse pattern of
Table~\ref{tab:simulation} does not transfer to verbatim negations because
(S1) fails in the field; V4's mechanism is vacuous under a calibrated
instrument because (S2) fails. The claim is re-scoped accordingly
(Prop.~\ref{prop:struct-equiv}, Remark~\ref{rem:boundary}).]}
(c) Lexicographic $\Pi$ vs.\ full $K(y)$ for repair routing.
(d) Mean-MMD vs.\ max-form $f_5$ on a single-negation adversarial suite: the
mean form's miss rate tends to $1$ as $m$ grows, by
Prop.~\ref{prop:dilution}.

\textbf{(4) Assumption tests.} Direct measurement of (S1) and (S2).
\emph{[Executed: both fail in the field; see \S\ref{sec:field}. This is the
protocol working as intended---the simulation's premises were falsifiable and
have been adjudicated.]}

\textbf{(5) Dependence diagnostics.} Empirical correlation matrix of
$(f_1,\dots,f_5)$; strong dependence motivates the safety-first
fixed-sequence procedure of Theorem~\ref{thm:familywise}.

\textbf{(6) Instrument audit.} Human-annotated decomposition benchmark for
$\varepsilon_D$ (claim-level recall, Clopper--Pearson interval);
edge-detection benchmark with adversarial paraphrase/negation for
$\varepsilon_{\mathrm{NLI}}$; abstention floor $c_{\min}$ reported with the
resulting abstention rate; all feeding the power certificate of
Prop.~\ref{prop:power}.

\textbf{(7) Drift monitoring.} E-process monitor state and recalibration
events logged; coverage audited within drift episodes, where the Regime-III
guarantee is exactly the testable object.

\textbf{(8) Registered next steps.} Scale-up ($n=1{,}000$/cell) with
template-family robustness; replication with an architecture-distinct
instrument (RoBERTa-large-MNLI); an organic-error corpus (model-generated
solutions with execution-labeled compositional errors) replacing template
insertion for the headline claim.
