% !TeX root = ../main.tex
\begin{thebibliography}{31}

\bibitem{aho1972} A.~V. Aho and T.~G. Peterson. A minimum-distance
error-correcting parser for context-free languages. \emph{SIAM J.\ Comput.},
1(4):305--312, 1972.

\bibitem{andoni2018} A.~Andoni, A.~Naor, and O.~Neiman. Snowflake
universality of Wasserstein spaces. \emph{Ann.\ Sci.\ \'Ec.\ Norm.\ Sup\'er.},
51(3):657--700, 2018.

\bibitem{angelopoulos2023} A.~N. Angelopoulos and S.~Bates. A gentle
introduction to conformal prediction and distribution-free uncertainty
quantification. \emph{arXiv:2107.07511}, 2023.

\bibitem{barber2021} R.~Foygel Barber, E.~J. Cand\`es, A.~Ramdas, and R.~J.
Tibshirani. The limits of distribution-free conditional predictive inference.
\emph{Information and Inference}, 10(2):455--482, 2021.

\bibitem{cherian2024} J.~Cherian, I.~Gibbs, and E.~Cand\`es. Large language
model validity via enhanced conformal prediction methods. \emph{NeurIPS},
2024.

\bibitem{cobbe2021} K.~Cobbe, V.~Kosaraju, M.~Bavarian, M.~Chen, H.~Jun,
L.~Kaiser, M.~Plappert, J.~Tworek, J.~Hilton, R.~Nakano, C.~Hesse, and
J.~Schulman. Training verifiers to solve math word problems.
\emph{arXiv:2110.14168}, 2021.

\bibitem{cofact2026} Z.~Hu, Z.~Zhang, Y.~Wang, L.~Rutkowski, and D.~Tao.
CoFact: Conformal factuality guarantees for language models under covariate
shift. \emph{ICLR}, 2026.

\bibitem{coles2001} S.~Coles. \emph{An Introduction to Statistical Modeling of
Extreme Values}. Springer, 2001.

\bibitem{cuturi2013} M.~Cuturi. Sinkhorn distances: lightspeed computation of
optimal transport. \emph{NeurIPS}, 2013.

\bibitem{fournier2015} N.~Fournier and A.~Guillin. On the rate of convergence
in Wasserstein distance of the empirical measure. \emph{Probab.\ Theory
Relat.\ Fields}, 162:707--738, 2015.

\bibitem{genevay2019} A.~Genevay, L.~Chizat, F.~Bach, M.~Cuturi, and
G.~Peyr\'e. Sample complexity of Sinkhorn divergences. \emph{AISTATS}, 2019.

\bibitem{ghrist2022} R.~Ghrist and H.~Riess. Cellular sheaves of lattices and
the Tarski Laplacian. \emph{Homology Homotopy Appl.}, 24(1):325--345, 2022.

\bibitem{gibbs2021} I.~Gibbs and E.~Cand\`es. Adaptive conformal inference
under distribution shift. \emph{NeurIPS}, 2021.

\bibitem{goemans1995} M.~X. Goemans and D.~P. Williamson. Improved
approximation algorithms for maximum cut and satisfiability problems using
semidefinite programming. \emph{J.\ ACM}, 42(6):1115--1145, 1995.

\bibitem{gretton2012} A.~Gretton, K.~M. Borgwardt, M.~J. Rasch,
B.~Sch\"olkopf, and A.~Smola. A kernel two-sample test. \emph{J.\ Mach.\
Learn.\ Res.}, 13:723--773, 2012.

\bibitem{guo2017} C.~Guo, G.~Pleiss, Y.~Sun, and K.~Q. Weinberger. On
calibration of modern neural networks. \emph{ICML}, PMLR 70:1321--1330, 2017.

\bibitem{hansen2019} J.~Hansen and R.~Ghrist. Toward a spectral theory of
cellular sheaves. \emph{J.\ Appl.\ Comput.\ Topol.}, 3:315--358, 2019.

\bibitem{he2021} P.~He, X.~Liu, J.~Gao, and W.~Chen. DeBERTa:
decoding-enhanced BERT with disentangled attention. \emph{ICLR}, 2021.

\bibitem{huntsman2024} S.~Huntsman, M.~Robinson, and L.~Huntsman. Prospects
for inconsistency detection using large language models and sheaves.
\emph{arXiv:2401.16713}, 2024.

\bibitem{inan2023} H.~Inan, K.~Upasani, J.~Chi, R.~Rungta, K.~Iyer, Y.~Mao,
M.~Tontchev, Q.~Hu, B.~Fuller, D.~Testuggine, and M.~Khabsa. Llama Guard:
LLM-based input-output safeguard for human-AI conversations.
\emph{arXiv:2312.06674}, 2023.

\bibitem{liang2023} P.~Liang et~al. Holistic evaluation of language models.
\emph{TMLR}, 2023.

\bibitem{lin2022} S.~Lin, J.~Hilton, and O.~Evans. TruthfulQA: measuring how
models mimic human falsehoods. \emph{ACL}, pages 3214--3252, 2022.

\bibitem{liu2019} Y.~Liu, M.~Ott, N.~Goyal, J.~Du, M.~Joshi, D.~Chen,
O.~Levy, M.~Lewis, L.~Zettlemoyer, and V.~Stoyanov. RoBERTa: a robustly
optimized BERT pretraining approach. \emph{arXiv:1907.11692}, 2019.

\bibitem{mohri2024} C.~Mohri and T.~Hashimoto. Language models with conformal
factuality guarantees. \emph{ICML}, 2024.

\bibitem{rebedea2023} T.~Rebedea, R.~Dinu, M.~Sreedhar, C.~Parisien, and
J.~Cohen. NeMo Guardrails: a toolkit for controllable and safe LLM
applications with programmable rails. \emph{EMNLP (System Demonstrations)},
2023.

\bibitem{reimers2019} N.~Reimers and I.~Gurevych. Sentence-BERT: sentence
embeddings using Siamese BERT-networks. \emph{EMNLP-IJCNLP}, pages
3982--3992, 2019.

\bibitem{tibshirani2019} R.~J. Tibshirani, R.~Foygel Barber, E.~Cand\`es, and
A.~Ramdas. Conformal prediction under covariate shift. \emph{NeurIPS}, 2019.

\bibitem{vovk2021} V.~Vovk. Testing randomness online. \emph{Statist.\ Sci.},
36(4):595--611, 2021.

\bibitem{williams2018} A.~Williams, N.~Nangia, and S.~R. Bowman. A
broad-coverage challenge corpus for sentence understanding through inference.
\emph{NAACL-HLT}, pages 1112--1122, 2018.

\bibitem{yadkori2024} Y.~Abbasi Yadkori, I.~Kuzborskij, A.~Gy\"orgy, and
C.~Szepesv\'ari. Mitigating LLM hallucinations via conformal abstention.
\emph{arXiv:2405.01563}, 2024.

\bibitem{zhou2023} J.~Zhou, T.~Lu, S.~Mishra, S.~Brahma, S.~Basu, Y.~Luan,
D.~Zhou, and L.~Hou. Instruction-following evaluation for large language
models. \emph{arXiv:2311.07911}, 2023.

\end{thebibliography}
